\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/introduction.tex
paper: another pass over intro and abstract

% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive.
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
In contrast to existing systems, \ProjectName{} exploits this observation and avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.

@@

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
\ProjectName{} contributes the first optimizer to plan data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
and express preferences in quality of data imputation versus running time. The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

@@

\subsection{Contributions}
This paper makes the following contributions:
\begin{itemize}
\item \textbf{Relational algebra with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\item \textbf{Proof of optimality:}
 We prove that our cost-model and planning algorithm produces a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive.
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.

@@

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics\todo{add more specific subarea qualifier} has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
\ProjectName{} contributes the first optimizer to plan data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
and express preferences in quality of data imputation versus running time. The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

@@

\subsection{Contributions}
This paper makes the following contributions:
\begin{itemize}
\item \textbf{Relational algebra with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\item \textbf{Proof of optimality:}
 We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-02-28 paper/main.tex
paper: minor fixes
paper: another pass over intro and abstract

\begin{abstract}
Missing values abound in data analysis and often present a serious challenge for usability of data. Users
are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
by  applying a potentially-expensive imputation strategy. We show that it is possible
to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
for each query. This allows users to immediately explore
their data, while the system picks the optimal placement of imputation operations. This placement is based
on the specific query issued and a parameter that the user selects for evaluating the tradeoff between data
quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
generates query plans that execute between \todo{add [low x - high x] range} as fast as imputing at the base
tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \todo{add SMAPE range}  
when compared to standard full-table imputation approach.
\end{abstract}

\input{introduction}
\input{examples}
\input{relatedwork}
\input{algorithm}
\input{experiments}
\input{conclusion}
\balance
\nocite{*}
\printbibliography

>>>>

\begin{abstract}
Missing values abound in data analysis and often present a serious challenge for usability of data. Users
are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
by  applying a potentially-expensive imputation strategy. We show that it is possible
to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
for each query. This allows users to immediately explore
their data, while the system picks the optimal placement of imputation operations. This placement is based
on the specific query issued and a user-provided parameter that trades off data
quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
generates query plans that execute between \todo{add [low x - high x] range} as fast as imputing at the base
tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \todo{add SMAPE range}  
when compared to standard full-table imputation approach.
\end{abstract}

\input{introduction}
\input{examples}
\input{relatedwork}
\input{algorithm}
\input{experiments}
\input{conclusion}
\balance
\printbibliography
